Quantifying the importance of a tree prior in Bayesian phylogenetics

### 2019-04-15 article.tex
Use @Giappo's workflow figure
Added @Giappo's figure

  \textbf{1. }
    BEAST2 is a popular Bayesian phylogenetics software tool,
    that takes an alignment of characters (usually nucleotides) 
    and an inference model to create a
    posterior of jointly-estimated phylogenies and model parameters. 
    An important ingredient of BEAST2 is the species tree prior, 
    due to which the tool can be used for the inference of 
    macroevolutionary processes underlying the species tree. 
    However, only models that allow for fast computation of 
    the prior probability are possible in practice. 
    This begs the question of how accurate the tree estimation is 
    when the real macroevolutionary processes are different 
    from those assumed in BEAST2. \\
  \textbf{2. }
    Here we present \verb;pirouette;, 
    a free, libre and open-source R package that assesses 
    the inference error BEAST2 makes based on a 
    phylogeny from any macroevolutionary diversification model. \\
  \textbf{3. }
    We describe \verb;pirouette;'s usage and the biological scientific
    question it can answer, including full examples. \\
  \textbf{4. }
    Last, we discuss the results obtained by the examples,
    their interpretation. \\
\end{abstract}

@@

BEAST [\cite{drummond2007beast}] is a Bayesian phylogenetics tool, 
that needs both data and priors to infer a posterior.
Of these priors, the tree prior describes how diversification occurs.
BEAST has built-in tree priors such as the Yule [\cite{yule}] and 
(constant-rate) birth-death [\cite{nee1994reconstructed}] model.
These simple tree priors are most commonly used, as these 
have sufficient biological complexity, while being computationally fast.
BEAST's successor, BEAST2 [\cite{bouckaert2014beast}],
has a package manager, that allows third-party users to extend existing 
functionalities.
By writing a BEAST2 package, novel diversification models can be added.
The key ingredient of such a package would be the likelihood equation
of a phylogeny for the novel diversification model.

@@

Even when a diversification model passes the procedure described above,
it is not necessarily used in Bayesian inference.
In Bayesian inference, the likelihood of a diversification 
model, is computed millions of times.
Due to this, effective but computationally expensive tree priors are non-viable.
Also, there may be no need to add such a tree prior to a Bayesian
inference framework: the more informative the data, 
the lower the influence the tree prior has.
Due to this, computionally faster tree priors are more commonly used.
However, it is unknown if it is valid to assume that tree prior choice is 
of low importance.

There have been multiple attempts to investigate the importance of tree
prior choice. For example, recently Sarver et al., [\cite{sarver2019choice}] 
show that the choice of tree prior does not 
substantially affect phylogenetic inferences of diversification rates.
Also recently, Duchene et al. [\cite{duchene2018phylodynamic}] released
a BEAST2 package to assess how well posterior predictive simulations
recover a given tree when using the standard diversification models.
These investigations show how current diversification models compare
to one another, but these do not help to assess the importance of a 
new tree prior.

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{workflow.png}
  \caption{
    \texttt{pirouette} pipeline.
    The pipeline starts from a phylogeny (1a) simulated by generative prior 
    $\mathit{P}$.
    The phylogeny is converted to an alignment (2a) using alignment model 
    $\mathit{A}$. 
    The user supplies one or more experiments.
    For each candidate experiment $\mathit{X}$ 
    (a tuple of inference model $\mathit{I}$ and condition $\mathit{C}$),
    if its condition $\mathit{C}$ is 
    satisfied (which can depend on the alignment), 
    it becomes a selected experiment $\mathit{X'}$, 
    to be used in the next step (3a).
    The inference models of the selected experiment use the alignment (2a) 
    to each create a Bayesian posterior of (parameter estimates and) 
    phylogenies (4a). 
    Each of the posteriors' trees is compared to the true phylogeny (1a) 
    using error measure $\mathit{E}$, 
    resulting in an error distribution (5a). 
    Optionally, a twin phylogeny (1b) can be generated from the original 
    phylogeny (1a) using twinning parameters $T$, where $T$ may depend on
    the experiments.
    From a twin phylogeny, the same intermediate steps are taken,
    resulting in a twin error distribution (5b)
  }
  \label{fig:pipeline}
\end{figure}

@@

The second step (Fig.~\ref{fig:pipeline}, 2a $\rightarrow$ 3a)
selects inference models from a set of experiments. 
We define an experiment $X$ as the combination of an inference model $I$
and the conditions $C$ to actually use it in the inference step.
There are multiple setups of condition $C$: 
either the attached inference model $I$ is the generative model and should always be selected,
or the attached inference model is a candidate model, that will be
be selected if it is the best model, or will be selected always (regardless
of it being the best). 
When setting $C$ to be attached to the generative model,
it is implied that attached inference model $I$ uses the 
same site and clock model used in the alignment simulation,
and $I$ uses the same tree prior as used in the creation of the given phylogeny. 
A 'best' model is defined as the inference model with
the highest evidence (a.k.a. marginal likelihood) for that inference model, 
given the alignment simulated in the previous step.
The evidence of an inference model is estimated by nested 
sampling [\cite{maturana2018model}], using the \verb;NS; BEAST2 package.
Scripted use of BEAST2 packages is only possible under under Linux and Mac.
Windows systems can do the model comparison for shorter DNA sequences
using the web interface of \verb;mcbette; [\cite{mcbette}].

>>>>

  \textbf{1. }
    BEAST2 is a popular Bayesian phylogenetics software tool,
    that takes an alignment of characters (usually nucleotides) 
    and an inference model to create a
    posterior of jointly-estimated phylogenies and model parameters. 
    An important ingredient of BEAST2 is the species tree prior, 
    due to which the tool can be used for the inference of 
    macroevolutionary processes underlying the species tree. 
    However, only models that allow for fast computation of 
    the prior probability are possible in practice. 
    This begs the question of how accurate the tree estimation is 
    when the real macroevolutionary processes are different 
    from those assumed in BEAST2. \\
  \textbf{2. }
    Here we present \verb;pirouette;, 
    a free, libre and open-source R package that assesses 
    the inference error BEAST2 makes based on a 
    phylogeny from any macroevolutionary diversification model. \\
  \textbf{3. }
    We describe \verb;pirouette;'s usage and the biological scientific
    question it can answer, including full examples. \\
  \textbf{4. }
    Last, we discuss the results obtained by the examples and their interpretation. \\
\end{abstract}

@@

BEAST [\cite{drummond2007beast}] is a Bayesian phylogenetics tool, 
that needs both data and priors to infer a posterior.
Of these priors, the tree prior describes how diversification occurs.
BEAST has built-in tree priors such as the Yule [\cite{yule}] and 
(constant-rate) birth-death [\cite{nee1994reconstructed}] models.
These simple tree priors are most commonly used, as these 
have sufficient biological complexity, while being computationally fast.
BEAST's successor, BEAST2 [\cite{bouckaert2014beast}],
has a package manager, that allows third-party users to extend existing 
functionalities.
By writing a BEAST2 package, novel diversification models can be added.
The key ingredient of such a package would be the likelihood equation
of a phylogeny for the novel diversification model.

@@

Even when a diversification model passes the procedure described above,
it is not necessarily used in Bayesian inference.
In these kind of procedures the likelihood of a diversification 
model can be computed millions of times.
Due to this, effective but computationally expensive tree priors are non-viable.
Also, there may be no need to add such a tree prior to a Bayesian
inference framework: the more informative the data, 
the lower the influence the tree prior has.
Due to this, computionally faster tree priors are more commonly used.
However, it is unknown if it is valid to assume that tree prior choice is 
of low importance.

There have been multiple attempts to investigate the importance of tree
prior choice. For example, recently Sarver et al., [\cite{sarver2019choice}] 
show that the choice of tree prior does not 
substantially affect phylogenetic inferences of diversification rates.
Also recently, Duchene et al. [\cite{duchene2018phylodynamic}] released
a BEAST2 package to assess how well posterior predictive simulations
recover a given tree when using the standard diversification models.
These investigations show how current diversification models compare
to one another, but they do not help to assess the importance of a 
new tree prior.

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{workflow_alt.png}
  \caption{
    \texttt{pirouette} pipeline.
    The pipeline starts from a phylogeny (1a) simulated by generative prior 
    $\mathit{P}$.
    The phylogeny is converted to an alignment (2a) using alignment model 
    $\mathit{A}$. 
    The user supplies one or more experiments.
    For each candidate experiment $\mathit{X}$ 
    (a tuple of inference model $\mathit{I}$ and condition $\mathit{C}$),
    if its condition $\mathit{C}$ is 
    satisfied (which can depend on the alignment), 
    it becomes a selected experiment $\mathit{X'}$, 
    to be used in the next step (3a).
    The inference models of the selected experiment use the alignment (2a) 
    to each create a Bayesian posterior of (parameter estimates and) 
    phylogenies (4a). 
    Each of the posteriors' trees is compared to the true phylogeny (1a) 
    using error measure $\mathit{E}$, 
    resulting in an error distribution (5a). 
    Optionally, a twin phylogeny (1b) can be generated from the original 
    phylogeny (1a) using twinning parameters $T$, where $T$ may depend on
    the experiments.
    From a twin phylogeny, the same intermediate steps are taken,
    resulting in a twin error distribution (5b)
  }
  \label{fig:pipeline}
\end{figure}

@@

The second step (Fig.~\ref{fig:pipeline}, 2a $\rightarrow$ 3a)
selects inference models from a set of experiments. 
We define an experiment $X$ as the combination of an inference model $I$
and the conditions $C$ to actually use it in the inference step.
There are multiple setups of $C$: 
either the attached inference model $I$ is the generative model and should always be selected,
or the attached inference model is a candidate model, that will be
be selected if it is the best model, or will be selected always (regardless
of it being the best). 
When setting $C$ to be attached to the generative model,
it is implied that attached inference model $I$ uses the 
same site and clock model used in the alignment simulation,
and $I$ uses the same tree prior as used in the creation of the given phylogeny. 
A 'best' model is defined as the inference model with
the highest evidence (a.k.a. marginal likelihood) for that inference model, 
given the alignment simulated in the previous step.
The evidence of an inference model is estimated by nested 
sampling [\cite{maturana2018model}], using the \verb;NS; BEAST2 package.
Scripted use of BEAST2 packages is only possible under under Linux and Mac.
Windows systems can do the model comparison for shorter DNA sequences
using the web interface of \verb;mcbette; [\cite{mcbette}].
